---
title: "STAT 440/540 Final Project Report"
author: "Group 01"
date: "2025-12-04"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
editor: visual
---

# Introduction

-   Question, context, why it matters
-   Dataset description, units, data dictionary (link or appendix)

## Methods

-   Data cleaning and preprocessing decisions (justify)
-   Modeling strategy
-   Assumptions for OLS and how you assessed them

## Results

-   Final model specification and fitted equation

```{r}
#| label: fit-final
# example refit; replace with your chosen model
# m_final <- lm(y ~ x1 + x2 + x3, data = dat)
```

```{r}
#| label:tbl-final
#| tbl-cap: "Final model coefficients with 95% CIs."
# broom::tidy(m_final, conf.int = TRUE) |>
#   knitr::kable(digits = 3)
```

## Diagnostics and Robustness

-   Residual plots, influence, multicollinearity (VIF), nonlinearity checks
-   Sensitivity (alternative functional forms, outlier handling, train/test split)

```{r}
#| label:fig-final-diags
#| fig-cap: "Diagnostics for the final model."
# performance::check_model(m_final)
```

## Discussion

-   Substantive takeaways; limitations; external validity
-   Ethical considerations and fairness (if relevant)

## Conclusion

-   What you learned; what you would do next with more time/data

## Reproducibility

```{r}
#| label: session-info
sessionInfo()
```

## Appendices

-   Extended tables/figures, data dictionary, additional models
