---
title: "STAT 440/540 Milestone 3: Multiple Regression & Refinement"
author: "Group 01"
date: "2025-10-30"
format:
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Introduction

-   Recap of EDA and initial insights guiding model building.

## Package Setup

```{r}
#| label: pkg-setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, broom, car, performance, interactions, splines, mgcv,
               kableExtra, ggplot2, patchwork)
set.seed(440540)
```

## Load Data

```{r}
#| label: load-data
dat <- readr::read_csv("../data/processed/dat_clean.csv")
```

## Candidate Models

We consider linear, interaction, polynomial, and spline specifications.

```{r}
#| label: fit-candidates
m_lin   <- lm(y ~ x1 + x2 + x3, data = dat)
m_int   <- lm(y ~ x1*x2 + x3, data = dat)
m_poly  <- lm(y ~ poly(x1,2) + x2 + x3, data = dat)
m_spline<- lm(y ~ splines::bs(x1, df=4) + x2 + x3, data = dat)
```

## Model Summaries

```{r}
#| label:tbl-candidates
#| tbl-cap: "Model fit summaries."
purrr::map_df(
  list(lin=m_lin, int=m_int, poly=m_poly, spline=m_spline),
  ~broom::glance(.x), .id="model"
) |>
  dplyr::select(model, r.squared, adj.r.squared, AIC, BIC, sigma) |>
  knitr::kable(digits=3)
```

## Coefficient Interpretation

```{r}
#| label:tbl-coefs
#| tbl-cap: "Coefficient estimates for interaction model."
broom::tidy(m_int, conf.int = TRUE) |>
  knitr::kable(digits=3)
```

## Diagnostics

```{r}
#| label:fig-diags
#| fig-cap: "Residual diagnostics for interaction model."
performance::check_model(m_int)
```

```{r}
#| label: vif
car::vif(m_int)
```

## Interaction Plot

```{r}
#| label:fig-interactions
#| fig-cap: "Interaction effect visualization."
interactions::interact_plot(m_int, pred = x1, modx = x2, centered = "all")
```

## Model Selection Rationale

-   Compare metrics and diagnostics; justify the preferred model.
-   Note limitations and trade-offs (bias–variance, interpretability).

## Reproducibility Snapshot

```{r}
#| label: session-info
sessionInfo()
```
